Bioconductor workflow for single-cell RNA sequencing: Normalization, dimensionality reduction, clustering, and lineage inference

### 2017-07-03 workflow/workflow.Rmd
Edits to text.

---
title: "Bioconductor workflow for single-cell RNA sequencing: Normalization, dimensionality reduction, clustering, and lineage inference"
author: 
  - name: Fanny Perraudeau
    affiliation: Graduate Group in Biostatistics, University of California, Berkeley
  - name: Davide Risso
    affiliation: Division of Biostatistics and Epidemiology, Department of Healthcare Policy and Research, Weill Cornell Medicine
  - name: Kelly N. Street
    affilitation: Graduate Group in Biostatistics, University of California, Berkeley
  - name: Elizabeth Purdom
    affiliation: Department of Statistics, University of California, Berkeley
  - name: Sandrine Dudoit
    affiliation: Graduate Group in Biostatistics, University of California, Berkeley; Division of Biostatistics and Department of Statistics, University of California, Berkeley

@@

@@

After assigning the cells to lineages and ordering them within lineages, we are interested in finding genes that have non-constant expression patterns over pseudotime. We can see global trends in expression by subsetting our original dataset to the most variable genes [SD: How many genes?] and displaying their expression values in a heatmap, with cells ordered by pseudotime [SD: Add figure ref.].

More formally, we use the robust local regression method loess to model in a flexible, non-linear manner the relationship between a gene's normalized expression measures and pseudotime. We then can test the null hypothesis of no change over time using the `gam` package. We implemented this approach for each gene and display the expression measures of the top 100 genes by p-value in the heatmap of Figure [SD: Add figure ref.].

@@

[SD: Add figure caption.]

```{r heatmapsignificant}
topgenes <- names(sort(gam.pval, decreasing = FALSE))[1:100]
heatdata <- y[rownames(se) %in% topgenes, order(t, na.last = NA)]
heatclus <- cl[order(t, na.last = NA)]
ce <- clusterExperiment(heatdata, heatclus, transformation = identity)
plotHeatmap(ce, clusterSamplesData = "orderSamplesValue", breaks = .99)
```

@@

[SD: Please check.]

@@

The authors are grateful to Professor John Ngai (Department of Molecular and Cell Biology, UC Berkeley) and his group members Dr. Russell B. Fletcher and Diya Das for motivating the research presented in this workflow and for valuable feedback on applications to biological data.

>>>>

---
title: "Bioconductor workflow for single-cell RNA sequencing: Normalization, dimensionality reduction, clustering, and lineage inference"
author: 
  - name: Fanny Perraudeau
    affiliation: Graduate Group in Biostatistics, University of California, Berkeley, Berkeley, CA
  - name: Davide Risso
    affiliation: Division of Biostatistics and Epidemiology, Department of Healthcare Policy and Research, Weill Cornell Medicine, New York, NY
  - name: Kelly N. Street
    affilitation: Graduate Group in Biostatistics, University of California, Berkeley, Berkeley, CA
  - name: Elizabeth Purdom
    affiliation: Department of Statistics, University of California, Berkeley, Berkeley, CA
  - name: Sandrine Dudoit
    affiliation: Graduate Group in Biostatistics, University of California, Berkeley; Division of Biostatistics and Department of Statistics, University of California, Berkeley, Berkeley, CA

@@

[SD: Refer to 3D version, used to create Figure 2.]

@@

After assigning the cells to lineages and ordering them within lineages, we are interested in finding genes that have non-constant expression patterns over pseudotime.

More formally, for each lineage, we use the robust local regression method loess to model in a flexible, non-linear manner the relationship between a gene's normalized expression measures and pseudotime. We then can test the null hypothesis of no change over time for each gene using the `gam` package. We implement this approach for the neuronal lineage and display the expression measures of the top 100 genes by p-value in the heatmap of Figure \@ref(fig:heatmapsignificant).

@@

```{r heatmapsignificant, fig.cap="Heatmap of the normalized expression measures for the 100 most significantly DE genes for the neuronal lineage, where rows correspond to genes and columns to cells ordered by pseudotime."}
topgenes <- names(sort(gam.pval, decreasing = FALSE))[1:100]
heatdata <- y[rownames(se) %in% topgenes, order(t, na.last = NA)]
heatclus <- cl[order(t, na.last = NA)]
ce <- clusterExperiment(heatdata, heatclus, transformation = identity)
plotHeatmap(ce, clusterSamplesData = "orderSamplesValue", breaks = .99)
```

@@

@@

The authors are grateful to Professor John Ngai (Department of Molecular and Cell Biology, UC Berkeley) and his group members Dr. Russell B. Fletcher and Diya Das for motivating the research presented in this workflow and for valuable feedback on applications to biological data. We would also like to thank Michael B. Cole for his contributions to `scone`.
